---
title: "BMI1"
output:
  word_document: default
  pdf_document: default
---
```{r}
library(readxl)
rm(list=ls())

D <- read.csv("C:/Users/vazol/Desktop/DTU/DTU/Introstat/bmi1/bmi1_data.csv",header = TRUE, sep=";",as.is=TRUE)
```

Descriptive analysis

a) Write a short description of the data. Which variables are included in the dataset?
Are the variables quantitative and/or categorized? (Categorized variables are only introduced in Chapter 8, but they are simply variables which divide the observations into categories/groups - e.g. three categories: low, medium, and high).How many observations are there? Are there any missing values? Remember to consult the extended description of the variables in Appendix 1 (p. 11).

```{r}
dim(D)
```
```{r}
names(D)

```
```{r}
head(D)

```
```{r}
tail(D)

```
```{r}
summary(D)
```
```{r}
str(D)
```

```{r}
D$bmi <- D$weight/(D$height/100)^2
```

b) Make a density histogram of the BMI scores. Use this histogram to describe the empirical distribution of the BMI scores. Is the empirical density symmetrical or skewed? Can a BMI score be negative? Is there much variation to be seen in the observations?


```{r}
mini=floor(min(D$bmi))
a=mini%%5
while (a!=0) {
  if(a==0){
    mini=mini
  }else{
    mini=mini-1
    a=mini%%5
  }
}

maxi=ceiling(max(D$bmi))
a=maxi%%5
while (a!=0) {
  if(a==0){
    maxi=maxi
  }else{
    maxi=maxi+1
    a=maxi%%5
  }
}


hist(D$bmi, xlab="BMI",main = "Histogram of: BMI",col=7,prob=TRUE,xlim=c(mini,maxi))
```

```{r}
Dfemale <- subset(D, gender == 0)
Dmale <- subset(D, gender == 1)
```

c) Make separate density histograms for the BMI scores of women and men, re-
spectively. Describe the empirical distributions of the BMI scores for men and
women using these histograms, like in the previous question. Does there seem
to be a gender difference in the distribution of the BMI scores (if so, describe the
difference)?

```{r}
mini=floor(min(Dfemale$bmi))
a=mini%%5
while (a!=0) {
  if(a==0){
    mini=mini
  }else{
    mini=mini-1
    a=mini%%5
  }
}

maxi=ceiling(max(Dfemale$bmi))
a=maxi%%5
while (a!=0) {
  if(a==0){
    maxi=maxi
  }else{
    maxi=maxi+1
    a=maxi%%5
  }
}


hist(Dfemale$bmi, xlab="BMI (female)", prob=TRUE,col=2,main ="Histogram of: BMI Female", xlim=c(mini,maxi))
```


```{r}
mini=floor(min(Dmale$bmi))
a=mini%%5
while (a!=0) {
  if(a==0){
    mini=mini
  }else{
    mini=mini-1
    a=mini%%5
  }
}

maxi=ceiling(max(Dmale$bmi))
a=maxi%%5
while (a!=0) {
  if(a==0){
    maxi=maxi
  }else{
    maxi=maxi+1
    a=maxi%%5
  }
}

hist(Dmale$bmi, xlab="BMI (male)", prob=TRUE,col=4,main ="Histogram of: BMI Male", xlim=c(mini,maxi))
```

d) Make a box plot of the BMI scores by gender. Use this plot to describe the em-
pirical distribution of the BMI scores for women and men. Are the distributions
symmetrical or skewed? Does there seem to be a difference between the distribu-
tions (if so, describe the difference)? Are there extreme observations/outliers?

```{r}
boxplot(Dfemale$bmi, Dmale$bmi,col = c(2,4), names=c("Female", "Male"), xlab="Gender", ylab="BMI")
```

e) Fill in the empty cells in the table above by computing the relevant summary
statistics for BMI, first for the full sample (both genders combined), then separately for women and men. Which additional information may be gained from
the table, compared to the box plot?

```{r}
table=matrix(c(sum(!is.na(D$bmi)),mean(D$bmi,na.rm=TRUE),var(D$bmi,na.rm=TRUE),sd(D$bmi,na.rm=TRUE),quantile(D$bmi,na.rm=TRUE,probs=0.25),quantile(D$bmi,na.rm=TRUE,probs=0.5),quantile(D$bmi,na.rm=TRUE,probs=0.75),sum(!is.na(Dfemale$bmi)),mean(Dfemale$bmi,na.rm=TRUE),var(Dfemale$bmi,na.rm=TRUE),sd(Dfemale$bmi, na.rm=TRUE),quantile(Dfemale$bmi,na.rm=TRUE,probs=0.25),quantile(Dfemale$bmi,na.rm=TRUE,probs=0.5),quantile(Dfemale$bmi,na.rm=TRUE,probs=0.75),sum(!is.na(Dmale$bmi)),mean(Dmale$bmi, na.rm=TRUE),var(Dmale$bmi, na.rm=TRUE),sd(Dmale$bmi, na.rm=TRUE),quantile(Dmale$bmi,na.rm=TRUE,probs=0.25),quantile(Dmale$bmi,na.rm=TRUE,probs=0.5),quantile(Dmale$bmi,na.rm=TRUE,probs=0.75)),ncol=7,byrow = TRUE)
#options("scipen"=100, "digits"=3)
colnames(table)=c('Number of obs.','Sample mean','Sample variance','Sample std. dev.','Lower quartile','Median','Upper Quartile')
rownames(table)=c('Everyone','Women','Men')
table=as.table(table)
table
table=as.data.frame.matrix(table)
```

Statistical analysis

f) Specify a statistical model for log-transformed BMI, making no distinction be-
tween men and women (see Remark 3.2). Estimate the parameters of the model
(mean and standard deviation). Perform model validation (see Chapter 3 and
Section 3.1.8). Since, in this case, confidence intervals and hypothesis tests in-
volve the distribution of an average, it might also be useful to include the central
limit theorem (Theorem 3.14) in the discussion.

```{r}
D$logbmi <- log(D$bmi)
qqnorm(D$logbmi)
qqline(D$logbmi)
alpha=mean(D$logbmi)
beta=sd(D$logbmi)
meanBMI=exp(alpha+beta**2/2)
sdBMI=exp(2*alpha+beta**2)*(exp(beta**2)-1)

plot(ecdf(D$logbmi), verticals = TRUE,main='ecdf:BMI')
xseq <- seq(0.9*min(D$logbmi), 1.1*max(D$logbmi), length.out = 100)
lines(xseq, pnorm(xseq, mean(D$logbmi), sd(D$logbmi)),col=2)

par(mfrow = c(3, 3))
for (i in 1:9){
xr <- rnorm(145)
qqnorm(xr, main="")
qqline(xr)
}

```

g) State the formula for a 95% confidence interval (CI) for the mean log-transformed
BMI score of the population (see Section 3.1.2). Insert values and calculate the
interval. Then, determine a 95% CI for the median BMI score of the population
(see Section 3.1.9).

```{r}
t95=qt(0.975,144)
meanBMI-(sqrt(sdBMI)/sqrt(length(D$logbmi)))*t95
meanBMI+(sqrt(sdBMI)/sqrt(length(D$logbmi)))*t95

t.test(exp(D$logbmi))
```


h) Perform a hypothesis test in order to investigate whether the mean log-transformed
BMI score is different from log(25). This can be done by testing the following
hypothesis, and corresponds to investigating whether the median BMI score is
different from 25:

```{r}
H0=log(25)
n=length(D$logbmi)
tobs=(mean(D$logbmi)-H0)/(sd(D$logbmi)/sqrt(n))
pvalue <- 2 * (1-pt(abs(tobs), df=n-1))
a=pt(abs(tobs), df=n-1)
pvalue

t.test(D$logbmi,mu=log(25))
```

i) Specify separate statistical models for log-transformed BMI for men and women.
Perform model validation for both models. Estimate the parameters of the models (mean and standard deviation for men and women, respectively).

```{r}
Dmale$logbmi <- log(Dmale$bmi)
qqnorm(Dmale$logbmi)
qqline(Dmale$logbmi)
alpham=mean(Dmale$logbmi)
betam=sd(Dmale$logbmi)
meanm=exp(alpham+betam**2/2)
sdm=exp(2*alpham+betam**2)*(exp(betam**2)-1)

plot(ecdf(Dmale$logbmi), verticals = TRUE,main='ecdf:BMI Male')
xseq <- seq(0.9*min(Dmale$logbmi), 1.1*max(Dmale$logbmi), length.out = 100)
lines(xseq, pnorm(xseq, mean(Dmale$logbmi), sd(Dmale$logbmi)),col=2)

par(mfrow = c(3, 3))
for (i in 1:9){
xr <- rnorm(73)
qqnorm(xr, main="")
qqline(xr)
}

```

```{r}
Dfemale$logbmi <- log(Dfemale$bmi)
qqnorm(Dfemale$logbmi)
qqline(Dfemale$logbmi)
alphaf=mean(Dfemale$logbmi)
betaf=sd(Dfemale$logbmi)
meanf=exp(alphaf+betaf**2/2)
sdf=exp(2*alphaf+betaf**2)*(exp(betaf**2)-1)

plot(ecdf(Dfemale$logbmi), verticals = TRUE,main='ecdf:BMI Female')
xseq <- seq(0.9*min(Dfemale$logbmi), 1.1*max(Dfemale$logbmi), length.out = 100)
lines(xseq, pnorm(xseq, mean(Dfemale$logbmi), sd(Dfemale$logbmi)),col=2)

par(mfrow = c(3, 3))
for (i in 1:9){
xr <- rnorm(74)
qqnorm(xr, main="")
qqline(xr)
}
```

j) Calculate 95% confidence intervals for the mean log-transformed BMI score for
women and men, respectively (se Section 3.1.2). Use these to determine 95%
confidence intervals for the median BMI score of women and men, respectively.
Fill in the table below with the confidence intervals for the two medians.

```{r}
t95f=qt(0.975,71)
t95m=qt(0.975,72)

table1=matrix(c(meanf-(sqrt(sdf)/sqrt(length(Dfemale$logbmi)))*t95f,meanm-(sqrt(sdm)/sqrt(length(Dmale$logbmi)))*t95m,meanf+(sqrt(sdf)/sqrt(length(Dfemale$logbmi)))*t95f,meanm+(sqrt(sdm)/sqrt(length(Dmale$logbmi)))*t95m),ncol=2,byrow = TRUE)
colnames(table1)=c('Lower bound of CI','Upper bound of CI')
rownames(table1)=c('Women','Men')
table1=as.table(table1)
table1
table1=as.data.frame.matrix(table1)
```

```{r}
Dfemale <- subset(D, gender == 0)
KI <- t.test(Dfemale$logbmi, conf.level=0.95)$conf.int
KI
exp(KI)
```

k) Perform a hypothesis test in order to investigate whether there is a difference
between the BMI of women and men. Specify the hypothesis as well as the significance level α, the formula for the test statistic, and the distribution of the test
statistic (remember the degrees of freedom). Insert relevant values and compute
the test statistic and p-value. Write a conclusion in words.

```{r}
mean1=mean(D$logbmi[D$gender==0])
mean2=mean(D$logbmi[D$gender==1])
H01=0
n1=length(D$logbmi[D$gender==0])
n2=length(D$logbmi[D$gender==1])
sd1=sd(D$logbmi[D$gender==0])
sd2=sd(D$logbmi[D$gender==1])
sd1=sd1**2
sd2=sd2**2
tobs1=((mean1-mean2)-H01)/sqrt(sd1/n1+sd2/n2)
v=((sd1/n1+sd2/n2)**2)/((sd1/n1)**2/(n1-1)+((sd2/n2)**2)/(n2-1))
pvalue1 <- 2 * (1-pt(abs(tobs1), df=v))
pvalue1
```

```{r}
t.test(D$logbmi[D$gender == 0], D$logbmi[D$gender == 1])
```


l) Comment on whether it was necessary to carry out the hypothesis test in the
previous question, or if the same conclusion could have been drawn from the
confidence intervals alone? (See Remark 3.59).

Correlation

m) State the formula for computing the correlation between BMI and weight. Insert
values and calculate the correlation (note, insert only numbers in the correlation
formula, i.e. three numbers). Furthermore, compute the remaining pairwise cor-
relations involving BMI, weight and fast food. Make pairwise scatter plots of
these variables. Assess whether the relation between the plots and the correlations is as you would expect.

```{r}
mean((D$weight-mean(D$weight))*(D$bmi-mean(D$bmi)))/(sd(D$weight)*sd(D$bmi))
cor(D$weight,D$bmi)
plot(x = D$weight,y = D$bmi,main = 'Correalations Weight-BMI',xlab = 'Weight',ylab = 'BMI',col='green')
plot(x = D$fastfood,y = D$bmi,col='blue',main = 'Correalations Fastfood-BMI',xlab = 'Fastfood',ylab = 'BMI')
plot(y = D$weight,x = D$fastfood,col='red',main = 'Correalations Weight-Fastfood',xlab = 'Weight',ylab = 'Fastfood')
```

```{r}
cor(D[,c("weight","fastfood","bmi")], use="pairwise.complete.obs")
```

