---
title: "R Notebook_fjord"
output:
  word_document: default
  html_notebook: default
---

a) Present a short descriptive analysis and summary of the data for the variables
t o t a l P , l o g c h l o r o p h y l , and t e m p . Include scatter plots of the log-transformed chlorophyl concentration against the two other variables, as well as histograms
and box plots of all three variables. Present a table containing summary statistics, which includes the number of observations, and the sample mean, standard deviation, median, and 0.25 and 0.75 quantiles for each variable. Furthermore, state which time period the observations cover.

```{r}
rm(list=ls())
D <- read.table("C:/Users/vazol/Desktop/DTU/DTU/Introstat/skivefjord2/skivefjord2_data.csv", header = TRUE, sep = ";")
```

```{r}
D$logchlorophyl <- log(D$chlorophyl)
```

```{r}
hist(D$logchlorophyl,xlab="LogChlorophyl",main ="Histogramof: LogChlorophyl",col=7,prob=TRUE)
```

```{r}
hist(D$temp,xlab="Temperature",main ="Histogramof: Temperature",col=2,prob=TRUE,xlim=c(-5,25))
```

```{r}
hist(D$totalP,xlab="totalP",main ="Histogramof: totalP",col=4,prob=TRUE)
```

```{r}
boxplot(D$logchlorophyl,xlab="LogChlorophyl",col = 7)
```

```{r}
boxplot(D$temp,xlab="Temperature",col = 2)
```

```{r}
boxplot(D$totalP,xlab="totalP",col = 4)
```

```{r}
table=matrix(c(sum(!is.na(D$logchlorophyl)),mean(D$logchlorophyl,na.rm=TRUE),sd(D$logchlorophyl,na.rm=TRUE),quantile(D$logchlorophyl,na.rm=TRUE,probs=0.25),quantile(D$logchlorophyl,na.rm=TRUE,probs=0.5),quantile(D$logchlorophyl,na.rm=TRUE,probs=0.75),sum(!is.na(D$totalP)),mean(D$totalP,na.rm=TRUE),sd(D$totalP,na.rm=TRUE),quantile(D$totalP,na.rm=TRUE,probs=0.25),quantile(D$totalP,na.rm=TRUE,probs=0.5),quantile(D$totalP,na.rm=TRUE,probs=0.75),sum(!is.na(D$temp)),mean(D$temp, na.rm=TRUE),sd(D$temp,na.rm=TRUE),quantile(D$temp,na.rm=TRUE,probs=0.25),quantile(D$temp,na.rm=TRUE,probs=0.5),quantile(D$temp,na.rm=TRUE,probs=0.75)),ncol=6,byrow = TRUE)
#options("scipen"=100, "digits"=3)
colnames(table)=c('Number of obs.','Sample mean','Sample std. dev.','Lower quartile','Median','Upper Quartile')
rownames(table)=c('Logchlorophyl','TotalP','Temp')
table=as.table(table)
table
table=as.data.frame.matrix(table)
```

```{r}
D_model <- D[1:234, ]
D_test <- D[235:240,]
```

c) Estimate the parameters of the model. These consist of the regression coefficients,
which we denote by β0, β1, β2, and the variance of the residuals, σ2. You may use
the following R code:

```{r}
fit <- lm(logchlorophyl ~ totalP + temp, data = D_model)
summary(fit)
```

d) Perform model validation with the purpose of assessing whether the model assumptions hold. Use the plots, which can be made using the R code below, as a
starting point for your assessment. (See section 6.4 on residual analysis).


```{r}
plot(fit$fitted.values, D_model$logbmi, xlab = "Fitted values",     
       ylab = "log(chlorophyl concentration)")

plot(D_model$totalP, fit$residuals, 
        xlab = "TotalP", ylab = "Residuals")

plot(D_model$temp, fit$residuals, 
        xlab = "Temp", ylab = "Residuals")

plot(fit$fitted.values, fit$residuals, xlab = "Fitted values", 
     ylab = "Residuals")

qqnorm(fit$residuals, ylab = "Residuals", xlab = "Z-scores", 
       main = "")
qqline(fit$residuals)
```

Ezt még utána sokkal részletesebben validálni kell a könyv és az első feladat alapján, esetleg még egy jó plot a fittedről a végén de nem itt hanem a h ban

e) State the formula for a 95% confidence interval for the age coefficient, here de-
noted by β1. (See Method 6.5). Insert numbers into the formula, and compute the
confidence interval. Use the R code below to check your result, and to determine
confidence intervals for the two other regression coefficients.

```{r}
beta1=fit$coefficient[2]
beta1+c(-1,1)*qt(0.975,231)*0.97795

```

```{r}
confint(fit, level = 0.95)
```

f) It is of interest whether β1 might be 0.001. Formulate the corresponding hypothesis. Use the significance level α = 0.05. State the formula for the relevant test
statistic (see Method 6.4), insert numbers, and compute the test statistic. State the
distribution of the test statistic (including the degrees of freedom), compute the
p-value, and write a conclusion.


```{r}
beta0.i=5
tobs=(beta1-beta0.i)/0.97795
p_value=2*(1-pt(abs(tobs),231))
p_value
```

g) Use backward selection to investigate whether the model can be reduced. (See
Example 6.13). Remember to estimate the model again, if it can be reduced. State
the final model, including estimates of its parameters.

Erről kérdezni, a 6.13 ban miért az x1nek nagyobb a p value értéke? Hiszen a confit nem a p value értékeket jeleníti meg

According to the Confidence interval and the p-values we can not reduce our model.


h) Use your final model from the previous question as a starting point. Determine predictions and 95% prediction intervals for the log-transformed chlorophyl concentration, for each of the six observations in the validation set (D _ t e s t ). See Example 6.8, Method 6.9 and the R code below. Compare the predictions to the observed log-chlorophyl concentrations for the six observations in the validation set and make an assessment of the prediction capabilities of the final model

```{r}
pred <- predict(fit, newdata = D_test, 
                interval = "prediction", level = 0.95)

cbind(year = D_test$year, month = D_test$month, 
      logchlorophyl = D_test$logchlorophyl, pred)
```


```{r}
conf <- predict(fit, newdata=D_test,
interval="confidence")
plot(D_test$logchlorophyl, pred[ ,"fit"], type="l",
xlab="x1", ylab="Prediction")
#lines(D_test$logbmi, conf[ ,"lwr"], col="green", lty=2)
#lines(D_test$logbmi, conf[ ,"upr"], col="green", lty=2)
lines(D_test$logchlorophyl, pred[ ,"lwr"], col="red")
lines(D_test$logchlorophyl, pred[ ,"upr"], col="red")
legend("topleft", c("Prediction","Confidence band","Prediction band"),
lty=c(1,2,2), col=c(1,3,2), cex=0.7)
```


